---
title: "`{pycnogrid}`: Flexible pycnophylactic interpolation to discrete global and local grid systems"
author:
  - name: "Christopher D. Higgins"
    orcid: "0000-0002-3551-7750"
    email: "cd.higgins@utoronto.ca"
    affiliations: "University of Toronto"
vignette: >
  %\VignetteIndexEntry{Getting started with `{pycnogrid}`}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
knitr:
  opts_chunk:
    collapse: true
    comment: '#>'
bibliography: references.bib
abstract: |
  Geographic data aggregated to administrative and statistical reporting units may not align with the geographic frame associated with particular research questions. Such units can also present the scale and zoning issues central to the modifiable areal unit problem (MAUP). In response, researchers and practitioners may turn to areal interpolation methods that can be used to transfer data from one geometry to another, with Tobler's pycnophylactic interpolation method distinguished by its emphasis on mass or volume preservation and spatial smoothing. Moreover, alternative hierarchical discrete global grid systems (DGGSs) that preserve area or shape or local grid systems present interesting opportunities for specifying alternative spatial support systems that have the potential to be more robust to the MAUP. Building on existing implementations of Tobler's pycnophylactic interpolation, the {pycnogrid} package offers a more flexible implementation of the algorithm that extends it beyond regular lattices to support a range of grid types, including the H3, A5, S2, and ISEA DGGSs as well as rasters and other local grids. Built on contemporary simple features and grounded in the foundations of spatial analysis, {pycnogrid} provides a tool for creating gridded, spatially smooth, and mass-preserving representations of aggregate extensive data for use in further analyses.

---

```{r}
#| label: setup
#| message: false
#| include: false
library(dplyr)
library(sf)
library(tmap)
library(pycnogrid)
```

## Introduction

Spatial data capturing the socioeconomic, demographic, environmental, health, travel, and other characteristics of different places has become increasingly abundant. However, much of these data are released in a spatially aggregated form, represented through administrative and statistical reporting units such as census tracts, postal codes, or traffic analysis zones. While these spatial units facilitate data collection, privacy protection, and dissemination, they may not align with the geographic frame associated with particular research questions or analyses. This mismatch lies at the heart of the modifiable areal unit problem (MAUP) [@openshaw1979], which recognizes that observed spatial patterns and statistical relationships may vary according to the zoning system used to aggregate data. This includes the scale effect, related to the size of geographic units, and the zoning effect related to how the polygons are drawn. In response, researchers and practitioners may look to methods for transferring or interpolating information from source zones to more standardized target geographies while preserving known source zone totals. At the same time, the development of area or shape preserving hierarchical discrete global grid systems (DGGSs) and alternative local grid options presents an opportunity to extend traditional interpolation methods to accommodate different grid types.

Among areal interpolation methods for extensive data, @tobler1979's pycnophylactic interpolation is distinguished by its emphasis on mass or volume preservation and spatial smoothing. Rather than assuming constant distributions of the source zone values across the target cells, pycnophylactic interpolation iteratively smooths the estimated target values while enforcing the preservation of the source zone totals in the system. With sufficiently fine target geometries, the result is a continuous spatial surface that preserves known aggregate quantities while reducing the abrupt discontinuities across zones introduced by administrative boundaries. In doing so, the method acknowledges that spatial aggregation obscures local variation and seeks to recover a *plausible* representation of the spatial phenomena at a finer alternative level of geography. The resulting surface embodies the intuition that nearby locations are expected to exhibit greater similarity than more distant locations [@tobler1970; @tobler2004].

Several implementations of Tobler's pycnophylactic interpolation exist for popular computing languages, including `{pycno}` [@pycno2025] for R and within the `{tobler}` package [@eliknaap2026] for Python, which is part of the larger Python Spatial Analysis Library (PySAL) [@sergiorey2026]. These existing implementations have largely followed Tobler's initial formulation of pycnophylactic interpolation from source zones to target raster cells. For example, the `pycno()` function from the `{pycno}` package converts source polygons to a regular grid and initializes a density surface by assigning a uniform density to all grid cells within each source zone. The density surface is then iteratively smoothed across neighbouring grid cells while correction steps ensure that the integrated target values within each source polygon remain equal to their original observed total. Built on the legacy `{sp}` framework, the package requires `SpatialPolygonsDataFrame` inputs and outputs a `SpatialGridDataFrame`. The PySAL `tobler` implementation follows a similar raster-based philosophy but performs smoothing through convolution-based operations on an array, with repeated correction steps used to preserve source-zone totals in the target cells.

Building on these foundations, the primary contribution of `{pycnogrid}` is not a new pycnophylactic algorithm for users of the R computing language, but rather a more flexible implementation that extends pycnophylactic interpolation beyond regular raster lattices to a range of DGGS spatial grids. Capitalizing on the contemporary simple features `{sf}` [@pebesma2018] library and the neighbour relationships and spatial weights matrices that underpin much of modern spatial analysis and econometrics [@anselin1988] (through the `{spdep}` [@bivand2022] and `{sfdep}` [@sfdep2024] packages), `{pycnogrid}` implements pycnophylactic interpolation as a generalized neighbourhood-based smoothing problem. Rather than restricting smoothing to a regular raster lattice, values are smoothed using the neighbourhood structure defined by the underlying grid, allowing the method to be applied consistently across grid systems including H3, A5, S2, and various ISEA apertures, as well as traditional rasters and other local grids. This flexibility enables users to select a grid system whose geometric properties are better aligned with the requirements of subsequent analyses.

## Example Workflow

The primary function in {pycnogrid} is `to_grid()`, which takes the following parameters:

```{{r}}
#| label: to_grid
#| echo: true
pycnogrid::to_grid(
  source,
  value_col,
  id_col = NULL,
  grid_type = c("h3", "a5", "s2", "isea3h", "isea4h", "raster", "hex"),
  resolution,
  cell_inclusion = c("intersect", "centroid"),
  cell_allocation = c("centroid", "area"),
  nb_order = 1,
  max_iter = 500,
  tolerance = 1e-4,
  include_self = TRUE,
  missing_policy = c("abort", "warn", "ignore")
) 
```

where

- `source` is the source {sf} polygon layer containing the totals to be interpolated. Given the geometry calculations performed by the tool, only inputs with projected coordinate reference systems are accepted.
- `value_col` is the column in `source` containing the count variable to be smoothed and preserved
- `id_col` is an optional column uniquely identifying each source polygon, if omitted, an internal identifier is created
- `grid_type` specifies the target grid system. Supported options are H3, A5, S2, ISEA grids with aperture-3 and 4, and raster-derived polygon cells
- `resolution` controls the size of the target grid cells; its interpretation depends on the selected grid type
- `cell_inclusion` defines how candidate grid cells are selected for interpolation. With `"intersect"`, cells are included if they intersect a source polygon. With `"centroid"`, cells are included only when their centroid falls inside a source polygon.
- `cell_allocation` defines how source totals are allocated to grid cells. With `"centroid"`, each grid cell is assigned to the source polygon containing its centroid. With `"area"`, values are allocated in proportion to the area of overlap between source polygons and grid cells.

::: {.callout-warning}
## Fractional area allocation

As of version 0.2.1, `cell_allocation = "area"` is labelled experimental. It preserves the overall or global represented total but may not reproduce individual source-zone totals when cells cross source boundaries as some drift can occur due to the areal allocation of values and weights. Use centroid allocation when strict source-zone preservation is required.
:::

- `nb_order` specifies the neighbourhood order used during smoothing. A value of 1 uses immediately adjacent cells, while larger values extend the smoothing neighbourhood outwards from a given cell.
- `max_iter` sets the maximum number of smoothing iterations. If set to 0, the function returns the initial allocation without iterative smoothing.
- `tolerance` defines the convergence threshold. Iteration stops when the relative change in estimated cell densities falls below this value.
- `include_self` controls whether each cell includes its own current value when calculating the neighbourhood mean during smoothing.
- `missing_policy` determines how the function handles source polygons that receive no target grid cells, which might arise due to a mismatch in source polygon sizes and target grid cell resolutions. "abort" stops with an error, "warn" returns a warning, and "ignore" proceeds silently.

Together, these parameters define three main stages of the workflow: construction of the target grid, allocation of source totals to grid cells, and iterative pycnophylactic smoothing. The grid parameters determine the spatial support of the output surface, the allocation parameters determine the initial mass-preserving estimate, and the smoothing parameters control how values are redistributed across neighbouring cells. Centroid-based allocation preserves the original source-zone totals with high accuracy while areal allocation preserves global totals but can result in some drift in source values (see the details below). Additional functions, such as `to_h3()` and `to_a5()`, are grid-specific wrappers around `to_grid()`. The properties of the different grid systems are discussed further below.

### Source Zone Data

The first part of a pycnophylactic workflow is to obtain more aggregate source data to be smoothed to a finer-resolution target grid. Pycnophylactic interpolation is designed for non-negative extensive variables -- quantities such as counts that can meaningfully be divided among smaller areas and whose total should be preserved. Due to the area and intersection calculations required for pycnophylactic smoothing, the source zones must be provided to the tool in a projected coordinate reference system. For an example case, 2020 population data was obtained for census tracts in New York City using the {tidycensus} [@tidycensus2026] package. A sub-sample of this data covering an area of Lower Manhattan is included in the package as `nyc_ct_small` (@fig-sample_population_count). The census tracts in this area contain a total of `r nyc_ct_small |> st_drop_geometry() |> summarize(sum(populationE)) |> prettyNum(big.mark = ",")` people.

```{r}
#| label: fig-sample_population_count
#| fig-cap: "Sample population count data for Lower Manhattan"
#| warning: false
#| echo: false
tm_shape(nyc_ct_small) + tm_fill(
  fill = "populationE",
  fill.scale = tm_scale_intervals(
    values = "viridis",
    style = "jenks",
    n = 6,
    label.format = tm_label_format(big.num.abbr = c("K" = 3))
  ),
  fill_alpha = .75,
  fill.legend = tm_legend(title = "Population Count", frame = FALSE)
) +
  #tm_basemap("CartoDB.PositronNoLabels", alpha = .5) +
  tm_shape(nyc_ct) + tm_fill(fill = "grey97", zindex = 3) +
  tm_layout(
    frame = FALSE,
    bg.color = "grey90"
  )
```

### Target Grid Systems

The choice of a target grid system is not merely a computational consideration as different grid systems prioritize different geometric and analytical properties, including equal-area representation, shape preservation, neighbourhood structure, hierarchical indexing, and scalable spatial computation (@tbl-grids).

```{r}
#| label: tbl-grids
#| tbl-cap: "Supported target grid systems and their principal geometric and hierarchical properties"
#| echo: false

tibble::tribble(
  ~grid_system, ~cell_geometry, ~area_property, ~extent_and_hierarchy,
  ~subdivision_or_aperture, ~neighbour_topology, ~analytical_implication,

  "H3",
  "Mostly hexagons, with twelve pentagons",
  "Approximately equal-area",
  "Global discrete grid with a nested hierarchy",
  "Aperture-7 hierarchy with alternating cell orientation across resolutions",
  "Usually six edge-neighbours; pentagons have five",
  "Strong indexing and neighbourhood structure, particularly useful for mobility, accessibility, and spatial aggregation",

  "ISEA3H, ISEA4H", # ISEA7H",
  "Mostly hexagons, with twelve pentagons on an icosahedral projection",
  "Equal-area",
  "Global discrete grid with a nested hierarchy",
  "Aperture-3 and Aperture-4 hierarchies, #, and Aperture-7 hierarchies",
  "Usually six edge-neighbours; pentagons have five",
  "Equal-area hexagonal option with relatively fine-grained hierarchical scaling between resolutions",

  #"ISEA4H",
  #"Mostly hexagons, with twelve pentagons on an icosahedral projection",
  #"Equal-area",
  #"Global discrete grid with a nested hierarchy",
  #"Aperture-4 hierarchy",
  #"Usually six edge-neighbours; pentagons have five",
  #"Equal-area hexagonal option with a different resolution progression and nesting structure from aperture-3 systems",

  #"ISEA7H",
  #"Mostly hexagons, with twelve pentagons on an icosahedral projection",
  #"Equal-area",
  #"Global discrete grid with a nested hierarchy",
  #"Aperture-7 hierarchy",
  #"Usually six edge-neighbours; pentagons have five",
  #"Equal-area hexagonal option with resolution scaling broadly comparable to H3 but without H3's specific indexing system",

  #"ISEA4/3H",
  #"Mostly hexagons, with twelve pentagons on an icosahedral projection",
  #"Equal-area",
  #"Global discrete grid with a nested hierarchy",
  #"Mixed aperture-4 and aperture-3 hierarchy",
  #"Usually six edge-neighbours; pentagons have five",
  #"Allows a mixed resolution progression where the scale change between successive levels follows the selected aperture sequence",
  
  "A5",
  "Equal-area pentagonal cells",
  "Equal-area",
  "Global discrete grid with a hierarchical structure",
  "Five-way initial refinement, followed by four-way refinement",
  "Predominantly five edge-neighbours",
  "Provides a global equal-area alternative with strong indexing, although its pentagonal geometry may produce more directional variation than hexagonal grids",

  "S2",
  "Quadrilateral cells projected from the faces of a cube",
  "Not equal-area",
  "Global discrete grid with a nested hierarchy",
  "Quadtree subdivision: each cell has four children",
  "Variable topology across cube-face boundaries",
  "Strong global indexing and web-mapping infrastructure, but cell areas vary substantially across locations",
  
  "Local Raster",
  "Rectangular cells in a projected CRS",
  "Equal-area only when constructed in an appropriate projected CRS",
  "Usually regional or local; hierarchy is not intrinsic",
  "Resolution specified directly in map units",
  "Four- or eight-neighbour structure, depending on rook or queen contiguity",
  "Simple and familiar benchmark, but geometric properties depend on the selected projection and resolution",
  
  "Local Hex",
  "Regular hexagonal cells in a projected CRS",
  "Equal-area only when constructed in an appropriate projected CRS",
  "Usually regional or local; hierarchy is not intrinsic",
  "Resolution specified directly in map units",
  "Usually six edge-neighbours in a complete tessellation",
  "Local hexagonal support with six symmetric first-order neighbours, but geometric properties depend on the selected projection and resolution"
) |>
  gt::gt() |>
  gt::cols_label(
    grid_system = "Grid System",
    cell_geometry = "Cell Geometry",
    area_property = "Area Property",
    extent_and_hierarchy = "Extent and Hierarchy",
    subdivision_or_aperture = "Subdivision / Aperture",
    neighbour_topology = "Neighbour Topology",
    analytical_implication = "Analytical Implication"
  ) |>
  gt::tab_options(
    table.width = gt::pct(100),
    table.font.size = 12
  )
```

Sample discrete global grids generated at different resolution levels to cover the census tracts for Lower Manhattan from @fig-sample_population_count are shown in @fig-sample_dggs_grids. Among the DGGSs, the H3 grid system is a global-scale spatial indexing system developed by Uber to support routing and mobility analytics. It is a hierarchical hexagonal (mostly -- twelve pentagonal cells are required to accommodate the topology of a spherical surface) DGGS built on an icosahedral projection of the Earth with 16 resolution levels. H3 offers scalable hierarchical spatial indexing as each cell at a given level of the hierarchy can be sub-divided into a set of 7 child sells, although the cell tesselations rotate slightly at each resolution level. Compared to raster cells, the hexagonal tesselation also offers consistent neighbourhood relationships and more isotropic traversal pathways between neighbouring cells with approximately equal distances to all first-order neighbours. Because H3 is built on a gnomonic projection, hexagons tend to preserve their shape in projected spatial analytical workflows. However, because of this projection, H3 cells distort globally and are not equal area. While they may be approximately equal at the urban scale, cells compared at the global scale can differ in area quite significantly. Support for H3 in `{pycnogrid}` is offered through the `{h3o}` [@h3o2025] package.

The Icosahedral Snyder Equal Area (ISEA) grid system partitions the globe into equal area hexagons with 31 supported resolution levels. Similar to H3, twelve pentagonal cells are required. In contrast to H3, where cells are approximately equal area, ISEA grids prioritize identical cell areas with more aligned nesting. This comes at the cost of shape preservation, with ISEA hexagons appearing more elongated than H3 when represented in conventional geographic or local projected coordinate systems. The ISEA grid supports several different apertures, including Aperture-3, 4, 7, and 4/3 mixed. At Aperture-3, parent cells divide into 3 child cells, while at Aperture-7 the ISEA grid behaves similar to H3 with each cell splitting into 7 across resolution levels. Support for ISEA grids is currently limited to Aperture-3 and 4 and offered through the `{hexify}` [@hexify2026] package. 

The A5 grid system is a recent DGGS that is explicitly designed to be equal area across the globe. Based on pentagons and 31 different resolution levels, A5 cells at a given resolution sub-divide into 5 child cells. The primary motivation behind A5 is spatial analysis as the equal area properties of the grid system greatly simplify comparisons of densities, rates, and other aggregated quantities over space. However, the subdivision of parent into child cells is not regular with cell shapes becoming more elongated at higher resolutions, meaning the distance between cell centroids is not uniform. With these slightly elongated pentagonal shapes, A5 cells are more anisotropic than H3 cells, although the extent to which impacts traversal pathways across the equal area cell topology has not yet studied. Support for the A5 grid system is offered through the `{a5R}` [@a5R2026] package.

S2 was developed by Google for global-scale spatial indexing and geometric computation. S2 begins by projecting the Earth onto the flat planes of a cube. Within a 31-level resolution hierarchy, each cell sub-divides into 4 child cells. While use of H3 and A5 might be motivated by neighbourhood regularity and equal-area representation, respectively, S2 emphasizes hierarchical spatial indexing and efficient spherical geometry operations. Consequently, the use of S2 within {pycnogrid} may be attractive when interpolated data are intended for integration with large-scale geospatial databases, distributed computing environments, or workflows that already rely on S2 indexing. Support for S2 in `{pycnogrid}` is offered through the `{s2}` [@s22025] package. 

```{r}
#| label: fig-sample_dggs_grids
#| fig-cap: "Sample DGGS hierarchies at different resolutions"
#| fig-height: 6
#| fig-width: 9
#| warning: false
#| echo: false
sample_dggs_grids <- bind_rows(
  
  h3_grids <- purrr::map_dfr(
    list(8, 9, 10),
    ~ pycnogrid:::prepare_target_cells(
      source_polys = nyc_ct_small,
      grid_type = "h3",
      resolution = .x,
      cell_inclusion = "intersect"
    ) |> mutate(grid_type = "h3", cell_resolution = .x, zorder = cell_resolution)
  ),
  
  isea3h_grids <- purrr::map_dfr(
    list(17, 18, 19),
    ~ pycnogrid:::prepare_target_cells(
      source_polys = nyc_ct_small,
      grid_type = "isea3h",
      resolution = .x,
      cell_inclusion = "intersect"
    ) |> mutate(grid_type = "isea3h", cell_resolution = .x, zorder = cell_resolution)
  ),
  
  isea4h_grids <- purrr::map_dfr(
    list(13, 14, 15),
    ~ pycnogrid:::prepare_target_cells(
      source_polys = nyc_ct_small,
      grid_type = "isea4h",
      resolution = .x,
      cell_inclusion = "intersect"
    ) |> mutate(grid_type = "isea4h", cell_resolution = .x, zorder = cell_resolution)
  ),
  
  a5_grids <- purrr::map_dfr(
    list(13, 14, 15),
    ~ pycnogrid:::prepare_target_cells(
      source_polys = nyc_ct_small,
      grid_type = "a5",
      resolution = .x,
      cell_inclusion = "intersect"
    ) |> mutate(grid_type = "a5", cell_resolution = .x, zorder = cell_resolution)
  ),
  
  s2_grids <- purrr::map_dfr(
    list(14, 15, 16),
    ~ pycnogrid:::prepare_target_cells(
      source_polys = nyc_ct_small,
      grid_type = "s2",
      resolution = .x,
      cell_inclusion = "intersect"
    ) |> mutate(grid_type = "s2", cell_resolution = .x, zorder = cell_resolution)
  )
) |>
  mutate(grid_type = forcats::fct_relevel(grid_type, "h3", "isea3h", "isea4h", "a5", "s2")) |> 
  arrange(desc(zorder))

tm_shape(sample_dggs_grids) +
  tm_lines(
    col = "cell_resolution",
    col.scale = tm_scale_categorical(),
    col.free = TRUE,
    col.legend = tm_legend(title = "resolution", frame = FALSE)
  ) +
  #tm_basemap("CartoDB.PositronNoLabels", alpha = .5) +
  tm_facets(by = "grid_type",
            ncol = 3,
            free.coords = FALSE) +
  tm_shape(nyc_ct) + tm_fill(fill = "grey97", zindex = 3) +
  tm_layout(
    panel.label.frame = FALSE,
    panel.label.bg = FALSE,
    frame = FALSE,
    bg.color = "grey90"
  )
```

Of the supported local grid types in @fig-sample_local_grids, raster grids are the most familiar, representing geographic space as a regular lattice of equally sized cells. Their simplicity has made them the dominant representation for continuous spatial phenomena and the traditional support for pycnophylactic interpolation. However, raster grids are inherently planar, which requires projection choices, and neighbourhood relationships depend on user-specified contiguity definitions (e.g., rook or queen adjacency). In terms of indexing, raster cells are not uniquely identified by a global index and any hierarchical relationships among cells of different resolutions would have to be manually specified. The neighbourhood structure is also anisotropic, with greater distances required to traverse cells diagonally than horizontally or vertically. Support for rasters in `{pycnogrid}` is offered through the `{terra}` [@terra2026] package.

Finally, users can also generate local regular hexagonal tessellations. Unlike the global DGGS options, these grids are defined within the coordinate reference system of the input data and have no intrinsic global index or hierarchical structure. Their area and shape properties therefore depend on selecting an appropriate projected CRS, while their alignment and resolution are specified directly in the map units of that CRS. Built on the {sf} [@pebesma2018] package, this option provides a familiar local hexagonal support for analyses focused on a particular region or study area.

```{r}
#| label: fig-sample_local_grids
#| fig-cap: "Sample local grid hierarchies at different resolutions"
#| fig-height: 3
#| fig-width: 9
#| message: false
#| echo: false
sample_local_grids <- dplyr::bind_rows(
  raster_grids <- purrr::map_dfr(
    list(1000, 500, 250),
    ~ pycnogrid:::prepare_target_cells(
      source_polys = nyc_ct_small,
      grid_type = "raster",
      resolution = .x,
      cell_inclusion = "intersect"
    ) |>
      mutate(
        grid_type = "raster",
        cell_resolution = .x,
        zorder = 1 / cell_resolution
      )
  ),
  hex_grids <- purrr::map_dfr(
    list(1000, 500, 250),
    ~ pycnogrid:::prepare_target_cells(
      source_polys = nyc_ct_small,
      grid_type = "hex",
      resolution = .x,
      cell_inclusion = "intersect"
    ) |>
      mutate(
        grid_type = "hex",
        cell_resolution = .x,
        zorder = 1 / cell_resolution
      )
  )
) |> mutate(grid_type = forcats::fct_relevel(grid_type, "raster", "hex")) |>
  arrange(desc(zorder))

tm_shape(sample_local_grids) +
  tm_lines(
    col = "cell_resolution",
    col.scale = tm_scale_categorical(),
    col.free = TRUE,
    col.legend = tm_legend(title = "resolution", frame = FALSE)
  ) +
  #tm_basemap("CartoDB.PositronNoLabels", alpha = .5) +
  tm_facets_wrap(by = "grid_type",
            #ncol = 2,
            free.coords = FALSE) +
  tm_shape(nyc_ct) + tm_fill(fill = "grey97", zindex = 3) +
  tm_layout(
    panel.label.frame = FALSE,
    panel.label.bg = FALSE,
    frame = FALSE,
    bg.color = "grey90"
  )
```


### Pycnophylactic Interpolation

With source data collected, the pycnophylactic smoothing can now be run. In this example case, an H3 grid at a resolution of 10 is used:

```{r}
#| label: run to_grid
pycno_nyc_ct_small <- nyc_ct_small |>
  pycnogrid::to_grid(
    value_col = populationE,
    grid_type = "h3",
    resolution = 10,
    cell_allocation = "centroid"
  )
```

The results of this interpolation are shown in @fig-pycno_nyc_ct_small below:

```{r}
#| label: fig-pycno_nyc_ct_small
#| fig-cap: "Census tract population counts interpolated to an H3 grid"
#| echo: false
tm_shape(pycno_nyc_ct_small) + tm_fill(
  fill = "pycno_populationE",
  fill.scale = tm_scale_intervals(
    values = "viridis",
    style = "jenks",
    n = 6,
    label.format = tm_label_format(big.num.abbr = c("K" = 3))
  ),
  fill_alpha = .75,
  fill.legend = tm_legend(title = "Interpolated \nPopulation", frame = FALSE)
) +
  tm_shape(nyc_ct_small) + tm_lines(col = "grey45") +
  #tm_basemap("CartoDB.PositronNoLabels", alpha = .5) +
  tm_shape(nyc_ct) + tm_fill(fill = "grey97", zindex = 3) +
  tm_layout(
    frame = FALSE,
    bg.color = "grey90"
  )
```

The results for other grid types can be seen in @fig-pycno_nyc_ct_small_resolutions:

```{r}
#| label: fig-pycno_nyc_ct_small_resolutions
#| fig-cap: "Census tract population counts interpolated to all four grid types"
#| fig-height: 8
#| fig-width: 10
#| echo: false
grid_list <- list("h3", "isea3h", "isea4h", "a5", "s2", "raster", "hex")
resolution_list <- list(10, 20, 16, 16, 16, 100, 100)

pycno_nyc_ct_small_resolutions <- purrr::map2_dfr(
  grid_list, resolution_list, ~ to_grid(
    source = nyc_ct_small,
    value_col = "populationE",
    grid_type = .x,
    resolution = .y
  ) |> 
    mutate(grid_type = .x, resolution = .y, facet_label = paste0(grid_type, " at resolution ", resolution))
)

tm_shape(pycno_nyc_ct_small_resolutions |> filter(grid_type != "h3" & grid_type != "isea4h" & grid_type != "hex")) + tm_fill(
  fill = "pycno_populationE",
  fill.scale = tm_scale_intervals(
    values = "viridis",
    style = "jenks",
    n = 6,
    label.format = tm_label_format(big.num.abbr = c("K" = 3))
  ),
  fill_alpha = .75,
  fill.legend = tm_legend(title = "Interpolated \nPopulation", frame = FALSE),
  fill.free = TRUE
) +
  tm_facets(by = "facet_label", ncol = 2,
                 free.coords = TRUE) +
  tm_shape(nyc_ct_small) + tm_lines(col = "grey45") +
  #tm_basemap("CartoDB.PositronNoLabels", alpha = .5) +
  tm_shape(nyc_ct) + tm_fill(fill = "grey97", zindex = 3) +
  tm_layout(panel.label.frame = FALSE, panel.label.bg = FALSE, frame = FALSE, bg.color = "grey90")
```

The summary statistics of the population variable for the different grid specifications is shown in @tbl-grid_descriptives. While the different grid types and resolutions necessarily produce different mean population values across the target cells, the total population spread over the target cells is intact.

```{r}
#| label: tbl-grid_descriptives
#| tbl-cap: "Population descriptive statistics for different grid types"
#| echo: false
pycno_nyc_ct_small_resolutions |> 
  st_drop_geometry() |> 
  group_by(grid_type) |> 
  summarize(
    resolution = min(resolution),
    cell_count = n(), 
    mean_pop = mean(pycno_populationE), 
    sum_pop = sum(pycno_populationE)
    ) |> 
  ungroup() |> 
  mutate(grid_type = forcats::fct_relevel(grid_type, "h3", "isea3h", "isea4h", "a5", "s2", "raster", "hex")) |> 
  arrange(grid_type) |> 
  gt::gt() |> 
  gt::cols_label(
    grid_type = "grid type",
    resolution = "grid resolution",
    cell_count = "grid cell count",
    mean_pop = "mean cell population",
    sum_pop = "total population"
  )
```

The main customizations for `to_grid()` involve defining additional options to guide the pycnophylactic interpolation. First, the cell inclusion and allocation criteria impact how target cells are generated and how source values are allocated to them.

- with `cell_inclusion = "intersect"` and `cell_allocation = "centroid"`, all intersecting grid cells remain in the target grid, but source values are assigned only to cells whose centroids fall within a source polygon. Target cells whose centroids fall outside the source polygons do not receive any initial allocation of the source values. Compared to area-based allocation, this simpler allocation method is more sensitive to the grid resolution and the placement of target cell centroids. Values associated with any source polygons that have no assigned target cells, such as small or narrow polygons, will be omitted from the system. When `missing_policy = "abort"`, this will cause the tool to abort. These are the default settings.

- with `cell_inclusion = "centroid"` and `cell_allocation = "centroid"`, target cells are both selected and assigned according to centroid location. Each retained cell is associated with the source polygon containing its centroid, producing the simplest and most discrete source-to-grid assignment with the strongest concentrations of source values within the target cells. However, as above, the centroid-based allocation approach is sensitive to grid resolution and alignment relative to the source polygons, risking truncated spatial support of the source layers within the target cell tessellation and the potential loss of source values if no cell centroids fall within the original polygon boundaries.

- with `cell_inclusion = "intersect"` and `cell_allocation = "area"`, all cells intersecting the source polygons are retained, and source totals are initially allocated according to the area of overlap between source polygons and target cells. This is the most geographically complete and resolution robust representation of the source layer as the data from all source-target intersections are retained. Partially covered edge cells receive only the share of the source value associated with their covered area. At present, the smoothing can result in some drift of source zone totals wherein the interpolated values do not match the original source zone totals for a given zone after re-aggregation, although the global source total is reproduced.

- with `cell_inclusion = "centroid"` and `cell_allocation = "area"`, only target cells whose centroids fall within the source layer are retained, but the retained cells receive source values according to their actual areas of overlap. While this avoids retaining cells that merely touch a source boundary, the resulting grid may not fully cover the source polygons, leading to a truncation in the spatial support offered by the target cells. The area-based allocation nevertheless helps to preserve source values in the system based on the areal overlap of source polygons and target cells. At present, the smoothing can result in some drift of source zone totals wherein the interpolated values do not match the original source zone totals for a given zone after re-aggregation, although the global source total is reproduced.

The smoothed results using the default and other cell inclusion and allocation settings are shown in @fig-pycno_nyc_ct_small_combinations below.

```{r}
#| label: fig-pycno_nyc_ct_small_combinations
#| fig-cap: "Interpolated population counts with varying inclusion and allocation parameters"
#| fig-height: 8
#| fig-width: 8
#| echo: false
#| warning: false
combinations <- tidyr::expand_grid(
  inclusion_list = list("intersect", "centroid"),
  allocation_list = list("centroid", "area")
)

pycno_nyc_ct_small_combinations <- purrr::map2_dfr(
  combinations |> purrr::pluck("inclusion_list"),
  combinations |> purrr::pluck("allocation_list"),
  ~ to_grid(
    source = nyc_ct_small,
    value_col = "populationE",
    id_col = id,
    grid_type = "h3",
    resolution = 10,
    cell_inclusion = .x,
    cell_allocation = .y,
    nb_order = 1
  ) |>
    mutate(
      inclusion = paste0("inclusion: ", .x),
      allocation = paste0("allocation: ", .y)
    )
) |>
  mutate(
    inclusion = forcats::fct_relevel(inclusion, "inclusion: intersect", "inclusion: centroid"),
    allocation = forcats::fct_relevel(allocation, "allocation: centroid", "allocation: area")
    )

tm_shape(pycno_nyc_ct_small_combinations) + tm_fill(
  fill = "pycno_populationE",
  fill.scale = tm_scale_intervals(
    values = "viridis",
    style = "jenks",
    n = 6,
    label.format = tm_label_format(big.num.abbr = c("K" = 3))
  ),
  fill_alpha = .75,
  fill.legend = tm_legend(title = "Interpolated \nPopulation", frame = FALSE)
) +
  tm_facets_grid(rows = "inclusion",
                 columns = "allocation",
                 free.coords = FALSE) +
  tm_shape(nyc_ct_small) + tm_lines(col = "grey45") +
  #tm_basemap("CartoDB.PositronNoLabels", alpha = .5) +
  tm_shape(nyc_ct) + tm_fill(fill = "grey97", zindex = 3) +
  tm_layout(panel.label.frame = FALSE, panel.label.bg = FALSE, frame = FALSE, bg.color = "grey90")
```

The second customization involves setting the number of neighbours through the `nb_order` parameter. Here the default is `1`, which results in the spatial weights matrix accounting for spatial relationships between a given target cell and its first-order neighbours with Queen contiguity. Increasing the order of the neighbourhood includes target cells from a larger neighbourhood which this has the effect of further smoothing out the spatial patterns of the interpolated values in the target cells. This can be seen in @fig-pycno_nyc_ct_small_nb_order.

```{r}
#| label: fig-pycno_nyc_ct_small_nb_order
#| fig-cap: "Interpolated population counts with increasing neighbourhood order"
#| fig-height: 8
#| fig-width: 8
#| echo: false
neighbours_list <- seq(1, 4, by = 1)

pycno_nyc_ct_small_neighbourss <- purrr::map_dfr(
  neighbours_list, ~ to_grid(
    source = nyc_ct_small,
    value_col = "populationE",
    grid_type = "h3",
    resolution = 10,
    nb_order = .x
  ) |> 
    mutate(nb_order = .x, facet_label = paste0("nb_order = ", nb_order))
)

tm_shape(pycno_nyc_ct_small_neighbourss) + tm_fill(
  fill = "pycno_populationE",
  fill.scale = tm_scale_intervals(
    values = "viridis",
    style = "jenks",
    n = 6,
    label.format = tm_label_format(big.num.abbr = c("K" = 3))
  ),
  fill_alpha = .75,
  fill.legend = tm_legend(title = "Interpolated \nPopulation", frame = FALSE),
  fill.free = FALSE
) +
  tm_facets(by = "facet_label",
            ncol = 2,
            free.coords = TRUE) +
  tm_shape(nyc_ct_small) + tm_lines(col = "grey45") +
  #tm_basemap("CartoDB.PositronNoLabels", alpha = .5) +
  tm_shape(nyc_ct) + tm_fill(fill = "grey97", zindex = 3) +
  tm_layout(panel.label.frame = FALSE, panel.label.bg = FALSE, frame = FALSE, bg.color = "grey90")
```

## Conclusion

`{pycnogrid}` provides a flexible implementation of pycnophylactic interpolation for redistributing polygon-based count data to discrete global and local grid systems. By expressing smoothing through neighbourhoods and spatial weights, the package extends the approach beyond regular raster lattices to H3, A5, S2, and ISEA grids, as well as raster and local hexagonal grids.

The package makes explicit several analytical choices that are often implicit in interpolation workflows. Users can control how target cells are included, how source values are allocated, the neighbourhood order used for smoothing, and whether cells contribute to their own neighbourhood means. With the default centroid-based allocation, each allocated target cell is assigned to one source zone, allowing the correction step to preserve represented source-zone totals under a discrete cell-ownership model. Fractional area allocation provides a more geographically complete treatment of source–target intersections, but it is experimental in version 0.2.1: it preserves the overall represented total but may not reproduce each individual source-zone total when target cells cross source boundaries.

The resulting surfaces should be understood as analytically useful, plausible estimates rather than observations at a finer geographic scale. Grid geometry, resolution, inclusion rules, allocation method, and neighbourhood structure can all affect the resulting spatial distribution and should be selected with reference to the intended analysis. Moreover, by representing source values over target cells, this grid-based method can extend or extrapolate source values beyond the original source polygon boundaries, particularly when the allocation method is based on cell centroids.

Future work could introduce an exact non-negative correction for fractional area allocation, ancillary-data weighting [e.g. @tapp2010], alternative smoothing operators, and support for additional area- or shape-preserving grid systems. In its current form, `{pycnogrid}` provides a practical framework for comparing smooth gridded representations across alternative spatial support systems while making their allocation assumptions and preservation guarantees explicit.

## References
